---
title: "Gapminder project"
author: "Abdullah Farouk"
date: '2017-09-15'
output: github_document
          
---

## What does Gapminder do?

Gapminder is a **fact tank** that aims to educate the world about global development. It hopes to do so with the use of statistical information deemed reliable. They collaborate with the UN, universities and various public agents to undertake their research. For more details on them see <http://www.gapminder.org/about-gapminder/>.


I initialize my exploration of the dataset with the following code:

```{r}
library(tidyverse)
library(gapminder)
```

To study the structure of the dataset and learn more about it's variables and their values, I do the following:

```{r}
str(gapminder)
head(gapminder)
tail(gapminder)
length(gapminder$country)
```


The gapminder dataset has information on three interesting variables, for 1704 countries from around the world. They are:
1. Life Expectancy
2. Population
3. GDP per capita

They provide some fascinating interactions for us to study.


### Brief Analysis of Life Expectancy Trends

To begin lets visualize the variation, accross time, in life expectancy using a bar chart. 

```{r}
hist(gapminder$lifeExp, main = 'Histogram of Life Expectancy')
```

We see that there are a lot of countries with high life expectancies. Has this always been the case?

```{r}

plot(gapminder$year, gapminder$lifeExp)

```

We notice that the average life span has been going up over time. Might this be because people around the world are simply richer today than they were before?

```{r}
plot(gapminder$year, gapminder$gdpPercap)
```


This does seem to be the case. Hence this could be one possible explanation of the trends we see in this variable. I hope to explore more such trends in the next class.

